Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-05-25 sports_analytics.Rmd
Updates.

In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a possession at any given point in the game. 
@lindsey1963investigation makes what is likely the most impactful such line of inquiry in baseball, leading to the analysis of all types of common strategies.

For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out. 
@lindsey1963investigation used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these two situations. 
His, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

Mathematically, we define $X$ to be a random variable indicating the number of runs (points) that a team will score in the remainder of a period, or game. 
Let $s$ be a tuple that encodes the *state* of a game. 
Then our task is to estimate:
$$
  \mathbb{E}[X | s] \,,
$$
for any state $s$.

In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball.

The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* and it is foundational in baseball analytics.

@@

```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
library(chess)
#Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() %>% plot()
```

>>>>

In many sports, the first step towards an analytical understanding of the game is the estimation of the expected value of a game state at any given point in a sporting contest.

Mathematically, we define $X$ to be a random variable indicating the the value of a particular game state for a team over some determined amount of time (e.g. remainder of game, quarter, period, or inning). Let $s$ be a tuple that encodes the *state* of a game. 
Then our task is to estimate:
$$
  \mathbb{E}[X | s] \,,
$$
for any state $s$.

For simplicity, we start with baseball due to the discrete nature of the sport.  In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball (25, if you include the absorbing state of 3 outs), and we are often interested in the number of runs that will be scored from some state until the end of the inning.  Thus in this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$.  The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix*, and it is foundational in baseball analytics.

Early work on this topic can be found in @lindsey1963investigation, which used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning.  This line of work led to work analyzing all types of common baseball strategies.  For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out.  Theirs, and many subsequent analyses [@tango2007book], conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

(Comment: Page 10 of this paper (https://www.jstor.org/stable/pdf/167996.pdf) actually have the expected values that were estimated. Do we want to include an example of this table? Might be cool to compare to like the numbers we got from open WAR.)

@@

```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
#library(chess)
#Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
#game() %>% plot()
```
